---
title: "Does College Pay Off?"
author: "Sangeeth Deleep Menon, Benoy Joseph, Rohit Biju, Bryan Samuel James"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 9
    fig_height: 6
subtitle: Modeling the Return on Investment of U.S. Higher Education Using the College
  Scorecard
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
set.seed(42)
```

---

# 1. Introduction

College tuition in the U.S. has climbed sharply over the last two decades, yet the financial payoff of a degree looks very different depending on where you go, what you study, where you live, and who you are. This project uses the U.S. Department of Education's **College Scorecard** to take a data-driven look at that question.

We define **Return on Investment (ROI)** as the relationship between what a student pays (tuition, fees, and net price after financial aid) and what they earn after graduation.

**Central Research Question:**
> *Which institutional characteristics and program features best predict whether a college degree delivers strong earnings relative to its cost, and how do these factors differ across institution types and fields of study?*

**Hypotheses:**

- **H1:** Net price meaningfully predicts median earnings 10 years after enrollment, even after controlling for selectivity.
- **H2:** ROI varies significantly by institution type; for-profit schools deliver weaker earnings relative to cost.
- **H3:** Field of study is a stronger earnings predictor than institutional characteristics.
- **H4:** Higher graduation rates predict better earnings even after controlling for selectivity.

---

# 2. Setup & Libraries

```{r libraries}
library(MASS)       # load before tidyverse to avoid masking dplyr::select
library(tidyverse)
library(corrplot)
library(caret)
library(pROC)
library(factoextra)
library(scales)
library(knitr)
library(kableExtra)
```

---

# 3. Data Acquisition & Preprocessing

## 3.1 Load Data

Data downloaded from the [U.S. Department of Education College Scorecard](https://collegescorecard.ed.gov/data/).

```{r load-data}
# Load institution-level data
# Download from: https://collegescorecard.ed.gov/data/
# File: Most Recent Institution-Level Data (CSV)
inst_raw <- read_csv(
  "data/Most-Recent-Cohorts-Institution.csv",
  na = c("", "NA", "PrivacySuppressed"),
  show_col_types = FALSE
)

# Load field-of-study data
# File: Most Recent Field of Study Data (CSV)
fos_raw <- read_csv(
  "data/Most-Recent-Cohorts-Field-of-Study.csv",
  na = c("", "NA", "PrivacySuppressed"),
  show_col_types = FALSE
)

cat("Institution-level rows:", nrow(inst_raw), "\n")
cat("Field-of-study rows:   ", nrow(fos_raw),  "\n")
```

## 3.2 Filter & Select Key Variables

```{r filter-inst}
inst <- inst_raw |>
  # Keep only predominantly bachelor's-degree-granting institutions
  filter(PREDDEG == 3) |>
  # Select key variables
  select(
    UNITID, OPEID, INSTNM, STABBR, REGION,
    CONTROL, PREDDEG, UGDS,
    ADM_RATE, SAT_AVG,
    NPT4_PUB, NPT4_PRIV,
    C150_4,
    MD_EARN_WNE_P10,
    DEBT_MDN,
    starts_with("PCIP")
  ) |>
  # Coerce numeric columns
  mutate(across(c(ADM_RATE, SAT_AVG, NPT4_PUB, NPT4_PRIV,
                  C150_4, MD_EARN_WNE_P10, DEBT_MDN, UGDS,
                  starts_with("PCIP")), as.numeric)) |>
  # Drop rows with suppressed earnings outcome
  filter(!is.na(MD_EARN_WNE_P10))

cat("Institutions after filtering:", nrow(inst), "\n")
```

## 3.3 Feature Engineering

```{r feature-engineering}
inst <- inst |>
  mutate(
    # Unified net price: use public rate for public schools, private for others
    net_price = case_when(
      CONTROL == 1 ~ NPT4_PUB,
      CONTROL %in% c(2, 3) ~ NPT4_PRIV,
      TRUE ~ NA_real_
    ),
    # ROI: earnings 10yr after entry divided by net price
    ROI = MD_EARN_WNE_P10 / net_price,
    # Binary ROI label for classification
    high_roi = ifelse(ROI > median(ROI, na.rm = TRUE), 1, 0),
    # STEM-heavy flag: sum of STEM-related PCIP field shares
    stem_share = rowSums(
      pick(PCIP11, PCIP14, PCIP15, PCIP26, PCIP27, PCIP40),
      na.rm = TRUE
    ),
    is_stem_heavy = as.integer(stem_share > 0.3),
    # Readable institution type label
    inst_type = case_when(
      CONTROL == 1 ~ "Public",
      CONTROL == 2 ~ "Private Nonprofit",
      CONTROL == 3 ~ "Private For-Profit",
      TRUE ~ NA_character_
    ),
    # Log-transform skewed variables for modeling
    log_ugds      = log1p(UGDS),
    log_net_price = log1p(net_price),
    log_earnings  = log1p(MD_EARN_WNE_P10)
  )

cat("Rows with valid ROI:", sum(!is.na(inst$ROI)), "\n")
```

## 3.4 Missingness Analysis

```{r missingness}
miss_summary <- inst |>
  select(net_price, ADM_RATE, SAT_AVG, C150_4,
         MD_EARN_WNE_P10, DEBT_MDN, ROI) |>
  summarise(across(everything(), ~mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  arrange(desc(pct_missing))

miss_summary |>
  kable(digits = 1, col.names = c("Variable", "% Missing"),
        caption = "Missing Data Summary for Key Variables") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r drop-high-missing}
# Drop variables with > 40% missing
vars_to_drop <- miss_summary |> filter(pct_missing > 40) |> pull(variable)
cat("Variables dropped (>40% missing):", paste(vars_to_drop, collapse = ", "), "\n")

# Complete-case analysis on core modeling variables
model_df <- inst |>
  filter(!is.na(net_price), !is.na(ADM_RATE), !is.na(SAT_AVG),
         !is.na(C150_4), !is.na(MD_EARN_WNE_P10), !is.na(ROI)) |>
  mutate(REGION = factor(REGION), inst_type = factor(inst_type))

cat("Complete cases for modeling:", nrow(model_df), "\n")
```

---

# 4. Exploratory Data Analysis

## 4.1 Distribution of Key Variables

```{r dist-plots, fig.height=8}
dist_data <- model_df |>
  select(inst_type, net_price, MD_EARN_WNE_P10, ROI) |>
  pivot_longer(-inst_type, names_to = "variable", values_to = "value") |>
  mutate(variable = recode(variable,
    net_price          = "Average Net Price ($)",
    MD_EARN_WNE_P10    = "Median Earnings 10yr ($)",
    ROI                = "ROI (Earnings / Net Price)"
  ))

ggplot(dist_data, aes(x = value, fill = inst_type)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  facet_grid(inst_type ~ variable, scales = "free") +
  scale_x_continuous(labels = label_comma()) +
  scale_fill_manual(values = c("Public" = "#2196F3",
                                "Private Nonprofit" = "#4CAF50",
                                "Private For-Profit" = "#E91E63")) +
  labs(title = "Distribution of Net Price, Earnings, and ROI by Institution Type",
       x = NULL, y = "Count", fill = "Institution Type") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 9))
```

## 4.2 Net Price vs. Median Earnings

```{r scatter-cost-earnings}
ggplot(model_df, aes(x = net_price, y = MD_EARN_WNE_P10, color = inst_type)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.2) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = dollar_format()) +
  scale_color_manual(values = c("Public" = "#2196F3",
                                 "Private Nonprofit" = "#4CAF50",
                                 "Private For-Profit" = "#E91E63")) +
  labs(title = "Net Price vs. Median Earnings 10 Years After Entry",
       subtitle = "By Institution Type — with per-group linear regression smoother",
       x = "Average Net Price (USD)", y = "Median Earnings 10yr After Entry (USD)",
       color = "Institution Type") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## 4.3 ROI by Institution Type

```{r roi-by-type}
ggplot(model_df, aes(x = inst_type, y = ROI, fill = inst_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Public" = "#2196F3",
                                "Private Nonprofit" = "#4CAF50",
                                "Private For-Profit" = "#E91E63")) +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  labs(title = "ROI Distribution by Institution Type",
       subtitle = "ROI = Median Earnings (10yr) / Average Net Price",
       x = "Institution Type", y = "ROI", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## 4.4 Geographic Analysis — Median ROI by Region

```{r geo-analysis}
region_labels <- c(
  "1" = "New England", "2" = "Mid East", "3" = "Great Lakes",
  "4" = "Plains", "5" = "Southeast", "6" = "Southwest",
  "7" = "Rocky Mtn", "8" = "Far West", "9" = "Outlying Areas"
)

model_df |>
  mutate(region_name = recode(as.character(REGION), !!!region_labels)) |>
  group_by(region_name) |>
  summarise(median_roi = median(ROI, na.rm = TRUE),
            n = n(), .groups = "drop") |>
  arrange(desc(median_roi)) |>
  ggplot(aes(x = fct_reorder(region_name, median_roi), y = median_roi, fill = median_roi)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = round(median_roi, 2)), hjust = -0.1, size = 3.5) +
  scale_fill_gradient(low = "#FFCDD2", high = "#1565C0") +
  coord_flip() +
  labs(title = "Median ROI by Census Region",
       x = NULL, y = "Median ROI", fill = "Median ROI") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## 4.5 Correlation Heatmap

```{r corr-heatmap, fig.height=7, fig.width=7}
corr_vars <- model_df |>
  select(MD_EARN_WNE_P10, net_price, ADM_RATE, SAT_AVG,
         C150_4, UGDS, DEBT_MDN, ROI) |>
  rename(Earnings_10yr = MD_EARN_WNE_P10,
         Net_Price     = net_price,
         Admit_Rate    = ADM_RATE,
         SAT_Score     = SAT_AVG,
         Grad_Rate     = C150_4,
         Enrollment    = UGDS,
         Median_Debt   = DEBT_MDN)

corr_matrix <- cor(corr_vars, use = "pairwise.complete.obs")

corrplot(corr_matrix,
         method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45,
         addCoef.col = "black", number.cex = 0.75,
         col = colorRampPalette(c("#E91E63", "white", "#1565C0"))(200),
         title = "\nCorrelation Matrix of Key Variables")
```

## 4.6 Earnings by Field of Study

```{r field-earnings, fig.height=8}
# PCIP field labels (CIP codes)
pcip_labels <- tribble(
  ~code,   ~field,
  "PCIP01", "Agriculture",
  "PCIP03", "Natural Resources",
  "PCIP04", "Architecture",
  "PCIP05", "Area Studies",
  "PCIP09", "Communication",
  "PCIP10", "Comm. Technologies",
  "PCIP11", "Computer Science",
  "PCIP12", "Personal Services",
  "PCIP13", "Education",
  "PCIP14", "Engineering",
  "PCIP15", "Engineering Tech",
  "PCIP16", "Foreign Languages",
  "PCIP19", "Family Sciences",
  "PCIP22", "Legal Studies",
  "PCIP23", "English",
  "PCIP24", "Liberal Arts",
  "PCIP25", "Library Science",
  "PCIP26", "Biology",
  "PCIP27", "Mathematics",
  "PCIP29", "Military Studies",
  "PCIP30", "Interdisciplinary",
  "PCIP31", "Leisure Studies",
  "PCIP38", "Philosophy",
  "PCIP39", "Theology",
  "PCIP40", "Physical Sciences",
  "PCIP41", "Science Tech",
  "PCIP42", "Psychology",
  "PCIP43", "Criminal Justice",
  "PCIP44", "Public Admin",
  "PCIP45", "Social Sciences",
  "PCIP46", "Construction",
  "PCIP47", "Mechanic/Repair",
  "PCIP48", "Precision Production",
  "PCIP49", "Transportation",
  "PCIP50", "Visual/Performing Arts",
  "PCIP51", "Health Professions",
  "PCIP52", "Business",
  "PCIP54", "History"
)

# Weighted average earnings by dominant field
field_earnings <- model_df |>
  select(MD_EARN_WNE_P10, starts_with("PCIP")) |>
  pivot_longer(-MD_EARN_WNE_P10, names_to = "code", values_to = "share") |>
  filter(!is.na(share), share > 0) |>
  left_join(pcip_labels, by = "code") |>
  group_by(field) |>
  summarise(
    weighted_earnings = weighted.mean(MD_EARN_WNE_P10, share, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  filter(!is.na(field), n >= 10) |>
  arrange(desc(weighted_earnings))

ggplot(field_earnings, aes(x = fct_reorder(field, weighted_earnings),
                            y = weighted_earnings, fill = weighted_earnings)) +
  geom_col(alpha = 0.85) +
  scale_y_continuous(labels = dollar_format()) +
  scale_fill_gradient(low = "#FFCDD2", high = "#1565C0", labels = dollar_format()) +
  coord_flip() +
  labs(title = "Weighted Average Earnings by Field of Study",
       subtitle = "Weighted by fraction of degrees awarded in each field",
       x = NULL, y = "Weighted Avg. Median Earnings (10yr)", fill = "Earnings") +
  theme_minimal(base_size = 10)
```

---

# 5. Hypothesis Testing

## 5.1 H1 — Does Net Price Predict Earnings After Controlling for Selectivity?

```{r h1-test}
# Base model: selectivity only
h1_base <- lm(MD_EARN_WNE_P10 ~ ADM_RATE + SAT_AVG, data = model_df)

# Full model: add net price
h1_full <- lm(MD_EARN_WNE_P10 ~ ADM_RATE + SAT_AVG + net_price, data = model_df)

summary(h1_full)
```

```{r h1-anova}
# Partial F-test: does adding net price significantly improve fit?
anova(h1_base, h1_full)
```

**Interpretation:** If the net_price coefficient is statistically significant (p < 0.05) and the ANOVA partial F-test shows a significant improvement, H1 is supported — net price predicts earnings even after accounting for selectivity.

## 5.2 H2 — Does ROI Differ Significantly by Institution Type?

```{r h2-test}
# One-way ANOVA
h2_aov <- aov(ROI ~ inst_type, data = model_df)
summary(h2_aov)
```

```{r h2-tukey}
# Post-hoc pairwise comparisons
as.data.frame(TukeyHSD(h2_aov)$inst_type) |>
  round(4) |>
  kable(caption = "Tukey HSD Post-hoc Comparisons for ROI by Institution Type") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Interpretation:** A significant ANOVA F-test (p < 0.05) supports H2. Tukey post-hoc tests identify which specific pairs of institution types differ significantly in ROI.

## 5.3 H3 — Field of Study vs. Institution Characteristics

```{r h3-test}
# Model 1: only institutional characteristics
h3_inst <- lm(MD_EARN_WNE_P10 ~ ADM_RATE + SAT_AVG + net_price +
                inst_type + UGDS + REGION,
              data = model_df)

# Model 2: add STEM indicator (proxy for field of study)
h3_field <- lm(MD_EARN_WNE_P10 ~ ADM_RATE + SAT_AVG + net_price +
                 inst_type + UGDS + REGION + stem_share + is_stem_heavy,
               data = model_df)

# Compare R²
cat("Institutional model R²:         ", round(summary(h3_inst)$r.squared, 4), "\n")
cat("Institutional + Field model R²: ", round(summary(h3_field)$r.squared, 4), "\n")

# Partial F-test
anova(h3_inst, h3_field)
```

**Interpretation:** If adding field-of-study variables (STEM share) substantially increases R² and the partial F-test is significant, H3 is supported — field matters beyond institution alone.

## 5.4 H4 — Graduation Rate Predicts Earnings After Controlling for Selectivity?

```{r h4-test}
h4_model <- lm(MD_EARN_WNE_P10 ~ ADM_RATE + SAT_AVG + C150_4, data = model_df)
summary(h4_model)
```

**Interpretation:** A significant positive coefficient on `C150_4` (p < 0.05) supports H4 — graduation rate predicts earnings even after controlling for institutional selectivity.

---

# 6. Linear Regression Modeling

## 6.1 Train/Test Split

```{r train-test-split}
set.seed(42)
train_idx <- createDataPartition(model_df$MD_EARN_WNE_P10, p = 0.8, list = FALSE)
train_df  <- model_df[train_idx, ]
test_df   <- model_df[-train_idx, ]

cat("Training set size:", nrow(train_df), "\n")
cat("Test set size:    ", nrow(test_df),  "\n")
```

## 6.2 Multiple Linear Regression

```{r linear-regression}
lm_full <- lm(
  MD_EARN_WNE_P10 ~ net_price + ADM_RATE + SAT_AVG +
    inst_type + C150_4 + log_ugds + REGION + stem_share,
  data = train_df
)

summary(lm_full)
```

## 6.3 Forward Stepwise Selection

```{r stepwise}
# Fit null and full models for stepwise bounds
lm_null <- lm(MD_EARN_WNE_P10 ~ 1, data = train_df)

lm_step <- stepAIC(lm_null,
                   scope = list(lower = lm_null, upper = lm_full),
                   direction = "forward",
                   trace = FALSE)

cat("Selected predictors:\n")
print(names(coef(lm_step))[-1])
```

## 6.4 Test Set Evaluation

```{r lm-evaluation}
# Predictions on held-out test set
preds_lm <- predict(lm_step, newdata = test_df)

rmse_lm <- sqrt(mean((test_df$MD_EARN_WNE_P10 - preds_lm)^2, na.rm = TRUE))
r2_lm   <- cor(test_df$MD_EARN_WNE_P10, preds_lm, use = "complete.obs")^2

cat(sprintf("Test RMSE: $%s\n", format(round(rmse_lm), big.mark = ",")))
cat(sprintf("Test R²:   %.4f\n", r2_lm))
```

## 6.5 Residual Diagnostics

```{r residual-diagnostics, fig.height=8}
par(mfrow = c(2, 2))
plot(lm_step, which = 1:4)
par(mfrow = c(1, 1))
```

## 6.6 K-Fold Cross-Validation

```{r cross-validation}
ctrl_cv <- trainControl(method = "cv", number = 10, verboseIter = FALSE)

cv_model <- train(
  MD_EARN_WNE_P10 ~ net_price + ADM_RATE + SAT_AVG +
    inst_type + C150_4 + log_ugds + REGION + stem_share,
  data      = model_df,
  method    = "lm",
  trControl = ctrl_cv,
  na.action = na.omit
)

cat("10-Fold CV Results:\n")
print(cv_model$results[, c("RMSE", "Rsquared", "MAE")])
```

---

# 7. ROI Classification (Logistic Regression)

## 7.1 Class Balance

```{r class-balance}
table(model_df$high_roi) |>
  as.data.frame() |>
  setNames(c("ROI Class", "Count")) |>
  mutate(`ROI Class` = ifelse(`ROI Class` == 1, "High ROI", "Low ROI")) |>
  kable(caption = "Distribution of Binary ROI Classes") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 7.2 Logistic Regression Model

```{r logistic-regression}
train_cls <- train_df |> filter(!is.na(high_roi))
test_cls  <- test_df  |> filter(!is.na(high_roi))

logit_model <- glm(
  high_roi ~ net_price + ADM_RATE + SAT_AVG +
    inst_type + C150_4 + log_ugds + REGION + stem_share,
  data   = train_cls,
  family = binomial
)

summary(logit_model)
```

## 7.3 Predictions & Confusion Matrix

```{r confusion-matrix}
probs_logit <- predict(logit_model, newdata = test_cls, type = "response")
preds_logit <- ifelse(probs_logit > 0.5, 1, 0)

cm <- confusionMatrix(
  factor(preds_logit, levels = c(0, 1)),
  factor(test_cls$high_roi, levels = c(0, 1)),
  positive = "1"
)
print(cm)
```

## 7.4 ROC Curve & AUC

```{r roc-curve}
roc_obj <- roc(test_cls$high_roi, probs_logit, quiet = TRUE)

ggroc(roc_obj, colour = "#1565C0", linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "gray50") +
  annotate("text", x = 0.3, y = 0.2,
           label = sprintf("AUC = %.3f", auc(roc_obj)),
           size = 5, color = "#1565C0") +
  labs(title = "ROC Curve — High vs. Low ROI Classification",
       x = "Specificity", y = "Sensitivity") +
  theme_minimal(base_size = 12)

cat(sprintf("AUC: %.4f\n", auc(roc_obj)))
```

---

# 8. Dimension Reduction & Clustering

## 8.1 PCA on Field-Share Variables

```{r pca}
pcip_data <- model_df |>
  select(starts_with("PCIP")) |>
  select(where(~mean(is.na(.)) < 0.4)) |>
  na.omit()

pca_result <- prcomp(pcip_data, center = TRUE, scale. = TRUE)

# Variance explained
fviz_eig(pca_result,
          addlabels = TRUE,
          barfill = "#1565C0",
          barcolor = "white",
          linecolor = "#E91E63") +
  labs(title = "PCA Scree Plot — Field-Share Variables",
       x = "Principal Component", y = "% Variance Explained")
```

```{r pca-biplot}
fviz_pca_var(pca_result,
             col.var = "contrib",
             gradient.cols = c("#FFCDD2", "#1565C0"),
             repel = TRUE,
             title = "PCA Variable Contributions (Field Shares)")
```

## 8.2 K-Means Clustering

```{r elbow-method}
# Build clustering feature matrix: PCA scores + cost/selectivity/outcome
pca_scores <- as.data.frame(pca_result$x[, 1:5])
inst_for_cluster <- model_df[complete.cases(model_df[, c("net_price", "SAT_AVG", "ROI")]), ]
n_cluster_rows <- min(nrow(pca_scores), nrow(inst_for_cluster))

cluster_features <- bind_cols(
  pca_scores[1:n_cluster_rows, ],
  inst_for_cluster[1:n_cluster_rows, c("net_price", "SAT_AVG", "ROI")]
) |> scale()

# Elbow method
fviz_nbclust(cluster_features, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow Method — Optimal Number of Clusters") +
  theme_minimal()
```

```{r kmeans}
set.seed(42)
km_fit <- kmeans(cluster_features, centers = 4, nstart = 25, iter.max = 100)

# Attach cluster labels to institutions
clustered_df <- inst_for_cluster[1:n_cluster_rows, ] |>
  mutate(cluster = factor(km_fit$cluster))

cat("Cluster sizes:\n")
print(table(clustered_df$cluster))
```

## 8.3 Cluster Visualization in PCA Space

```{r cluster-viz}
fviz_cluster(km_fit,
             data = cluster_features,
             geom = "point",
             ellipse.type = "convex",
             palette = c("#2196F3", "#4CAF50", "#E91E63", "#FF9800"),
             ggtheme = theme_minimal()) +
  labs(title = "K-Means Clusters Visualized in PCA Space")
```

## 8.4 Cluster Profiles

```{r cluster-profiles}
cluster_profiles <- clustered_df |>
  group_by(cluster) |>
  summarise(
    n           = n(),
    median_roi  = round(median(ROI, na.rm = TRUE), 2),
    median_earn = round(median(MD_EARN_WNE_P10, na.rm = TRUE)),
    median_cost = round(median(net_price, na.rm = TRUE)),
    median_sat  = round(median(SAT_AVG, na.rm = TRUE)),
    pct_public  = round(mean(CONTROL == 1, na.rm = TRUE) * 100, 1),
    pct_forprofit = round(mean(CONTROL == 3, na.rm = TRUE) * 100, 1),
    .groups = "drop"
  )

cluster_profiles |>
  kable(
    col.names = c("Cluster", "N", "Median ROI", "Median Earnings ($)",
                  "Median Net Price ($)", "Median SAT", "% Public", "% For-Profit"),
    caption   = "Cluster Profiles: Cost, Selectivity, and Outcomes",
    format.args = list(big.mark = ",")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r cluster-top-schools}
# Top schools by ROI in each cluster
clustered_df |>
  group_by(cluster) |>
  slice_max(ROI, n = 3) |>
  select(cluster, INSTNM, inst_type, ROI, MD_EARN_WNE_P10, net_price) |>
  mutate(across(c(ROI), round, 2),
         across(c(MD_EARN_WNE_P10, net_price), ~dollar(round(.)))) |>
  kable(col.names = c("Cluster", "Institution", "Type", "ROI",
                       "Earnings (10yr)", "Net Price"),
        caption = "Top 3 Schools by ROI Within Each Cluster") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# 9. Discussion & Conclusions

## 9.1 Hypothesis Summary

```{r hypothesis-summary}
tribble(
  ~Hypothesis, ~Prediction, ~Result,
  "H1", "Net price predicts earnings after controlling for selectivity",
  "See Section 5.1 — coefficient on net_price in full model",
  "H2", "For-profit schools deliver lower ROI than public/nonprofit",
  "See Section 5.2 — ANOVA and Tukey post-hoc results",
  "H3", "Field of study is a stronger predictor than institution characteristics",
  "See Section 5.3 — R² comparison between models",
  "H4", "Higher graduation rates → better earnings after controlling for selectivity",
  "See Section 5.4 — coefficient on C150_4"
) |>
  kable(caption = "Hypothesis Testing Summary") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 9.2 Key Findings

- **ROI varies substantially** by institution type, with for-profit schools generally clustering at lower earnings relative to cost (consistent with preliminary EDA).
- **Net price is not always a proxy for quality** — after controlling for selectivity, additional cost does not uniformly translate to higher earnings.
- **Field of study drives significant variation** in outcomes, supporting the view that what you study matters alongside where you study.
- **Graduation rates signal institutional quality** beyond selectivity alone, suggesting that student support structures have measurable economic value.

## 9.3 Limitations

- Data is at the institution level; individual-level variation within schools cannot be captured.
- Earnings at 10 years may reflect cohort-specific economic conditions, not just school quality.
- Suppressed values for small programs introduce systematic missingness.
- Causality cannot be inferred — selection effects remain a fundamental challenge.

## 9.4 Policy Implications

These findings can help prospective students compare programs on a cost-adjusted basis, and encourage policymakers to consider ROI metrics (not just prestige) when designing federal aid policy.

---

# 10. References

- U.S. Department of Education. (2025). *College Scorecard Data*. https://collegescorecard.ed.gov/data/
- U.S. Department of Education. (2025). *College Scorecard Technical Documentation*. https://collegescorecard.ed.gov/assets/InstitutionDataDocumentation.pdf
- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning* (2nd ed.). Springer.
- Bruce, P., Bruce, A., & Gedeck, P. (2020). *Practical Statistics for Data Scientists* (2nd ed.). O'Reilly.
- Wickham, H., & Grolemund, G. (2023). *R for Data Science* (2nd ed.). O'Reilly. https://r4ds.hadley.nz/
- Aden-Buie, G. (2025). *collegeScorecard: US College Scorecard Data* (R package). CRAN. https://cran.r-project.org/package=collegeScorecard

---

*DS5110 — Data Management and Processing | Northeastern University | Spring 2026*
